---
title: "Getting Started with HistoneMod"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with HistoneMod}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
if (requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(".", export_all = FALSE, quiet = TRUE)
} else {
  library(HistoneMod)
}
```

## Overview

`HistoneMod` provides two complementary interfaces:

- an interactive Shiny application for end users;
- exported helper functions for validation, preprocessing, and plotting in
  scripted workflows.

This vignette shows both usage styles and gives a minimal reproducible example
that can be adapted to real Skyline-like peptide tables.

## Installation

Install the package from GitHub with dependencies enabled so required packages
are installed automatically:

```r
install.packages("remotes")
remotes::install_github("jiehua1995/HistoneMod", dependencies = TRUE)
```

After installation, you can optionally verify that runtime dependencies are
available:

```{r}
depends_check()
```

## Launch the Shiny application

The main entry point is `runHistoneMod()`:

```r
runHistoneMod()
```

If you want the application object itself, use:

```r
app <- histonemod_app()
shiny::runApp(app)
```

## Input tables

`HistoneMod` expects two CSV files:

- an `MS1` table with columns such as `Protein.Name`, `Peptide.Note`,
  `Replicate.Name`, `Total.Area.MS1`, and `Isotope.Label.Type`;
- a `Sample` table with columns such as `Replicate.Name`, `Group`, and
  `Replicate.No`.

The validation helpers can be used before launching the app or inside a custom
pipeline.

## Minimal reproducible example

The following synthetic example keeps the same column names that the package
expects in real input files.

```{r}
sample_tbl <- data.frame(
  Replicate.Name = c("Ctrl_1", "Ctrl_2", "Treat_1", "Treat_2"),
  Group = c("Ctrl", "Ctrl", "Treat", "Treat"),
  Replicate.No = c(1, 2, 1, 2)
)

ms1_tbl <- data.frame(
  Protein.Name = rep("H3_3-8", 12),
  Peptide.Note = rep(
    c("H3_3_8_K4_me1", "H3_3_8_K9_ac", "H3_3_8_K14_un"),
    times = 4
  ),
  Replicate.Name = rep(sample_tbl$Replicate.Name, each = 3),
  Total.Area.MS1 = c(
    40, 35, 25,
    38, 37, 25,
    20, 55, 25,
    18, 57, 25
  ),
  Isotope.Label.Type = "light"
)

sample_tbl
ms1_tbl
```

## Validate CSV inputs

The validation helpers operate on file paths, so a common pattern is to write or
receive CSV files and validate them before analysis.

```{r}
ms1_path <- tempfile(fileext = ".csv")
sample_path <- tempfile(fileext = ".csv")

write.csv(ms1_tbl, ms1_path, row.names = FALSE)
write.csv(sample_tbl, sample_path, row.names = FALSE)

validate_ms1_file(ms1_path)
validate_sample_file(sample_path)
```

## Compute relative peptide percentages

`percentage_calculation()` merges the two tables, filters to light peptides, and
computes relative abundance percentages within each protein and replicate.

```{r}
processed <- percentage_calculation(
  ms1 = ms1_tbl,
  sample = sample_tbl,
  exclude_un = TRUE
)

processed
```

You can also filter specific peptides or samples:

```{r}
percentage_calculation(
  ms1 = ms1_tbl,
  sample = sample_tbl,
  exclude_un = TRUE,
  selected_peptides = "H3_3_8_K9_ac",
  selected_samples = c("Ctrl_1", "Treat_1")
)
```

## PCA plot

```{r fig.width=7, fig.height=5}
plot_pca(processed, show_ellipse = FALSE)
```

## Heatmap

```{r fig.width=7, fig.height=5}
hm <- plot_heatmap(processed, cluster_rows = FALSE, cluster_cols = FALSE)
grid::grid.newpage()
grid::grid.draw(hm$gtable)
```

## Single-peptide barplot

```{r fig.width=7, fig.height=5}
plot_barplot_single(
  processed,
  protein_name = "H3_3-8",
  peptide_name = "H3_3_8_K9_ac",
  add_signif = FALSE
)
```

## Version helpers

`HistoneMod` also exports a small helper for normalizing release tags and a
function for checking the latest GitHub release version.

```{r}
normalize_version("v0.5.2")
```

The GitHub lookup is typically only needed in interactive workflows, because it
depends on network access:

```r
github_latest_release_version("jiehua1995/HistoneMod")
```

## Next steps

For exploratory analysis, most users should start with the Shiny interface. For
automated workflows, the exported validation, preprocessing, and plotting
functions provide a scriptable path that uses the same package logic.
